A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-11-08 thesis/thesis.tex
fix typos, add bsearch, handle block plymorphism

\begin{abstract}
In this work, we introduce a novel linearizable wait-free queue implementation.
Linearizability and lock-freedom are standard requirements for designing shared data structures. To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case called the CAS Retry Problem. We show that our algorithm avoids this problem and has worst case running time better than prior lock-free queues. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm, is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
\end{abstract}

\tableofcontents
\pagebreak
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure to work with. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can make progress at the time the data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored. \texttt{Dequeue} removes and returns the first element in the queue. If the queue is empty it returns \texttt{null}. In a concurrent version of the queue data structure, operations do not happen one at a time. The question that may arise is ``What properties matter for implementation of a shared data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes. Our model is an asynchronous shared-memory distributed system of $p$ processes. Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy on how powerful objects are to solve the consensus problem among processes. Objects like LL/SC, CAS can be used to reach consensus among any number of processors. A CAS object provides an atomic Compare \& Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}, otherwise it returns \nf{false}.  In this work we use Compare \& Swap (\nf{CAS}) objects to synchronize among  processes.

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/bad}
  \caption{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable. \texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occurr before termination of \texttt{Enqueue(A)} and return \nf{null} or can occurr after termination of \texttt{Enqueue(A)} and return \texttt{A}.}
\end{figure}

There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom. An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps. We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates. A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm there might be an operation that takes infinitely many steps but never terminates.

We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps. We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized. Processes can do queries to get information about the linearization stored in the root. We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated after at most 2 \nf{CAS} invocations. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized. \nf{CAS} instructions cost more than other instructions for processor. Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case. We use unbounded memory space in our queue, we present a way to reduce the memory size to the total number of operations. We do not address garbage collection but we describe a way we think it can be handled.

\paragraph{Thesis outline} The rest of the thesis is organized as follows. Chapter 2 gives an outline of related work done in the area and the motivation of our implementation. Previous lock-free queues and their common problem is presented in Section 2.1. In Section 2.2 we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects. Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.

In Chapter 3 we introduce a poly-logarithmic step wait-free queue. In Section 3.1 we give a high-level description of our implementation. We also talk about the motivation and requirements in our design to achieve poly-log time. Section 3.2 contains the algorithm itself.

@@

Chapter 5 is devoted to performance analysis for our implementation. We analyze the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue. In the end we prove our queue is wait-free.

@@

\pagebreak
\section{Related Work}
In this section, we look at previous lock-free queues. In the worst-case, step complexity of all the lock-free queues, includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or dequeue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.4]{pics/msqueue}
  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. In the first diagram the first element has been dequeued. Red arrows show new pointers and gray dashed arrows show the old pointers.}
\end{figure}

@@

Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue. Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket. Each basket has a time interval in which all its nodes' enqueue operations overlap. Since the operations in a basket are concurrent, we can order them in any way. Enqueues in a basket try to find their order in the basket one by one by using \texttt{CAS} operations. However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption{Baskets queue idea. There is a time that all operations in a basket were running concurrently, but only one has succeeded to do \texttt{CAS}. To order the operations in a basket, the mechanism in the algorithm for processes is to \texttt{CAS} again. The successful process will be the next one in the basket and so on.}
\end{figure}
% \bf{TODO: make small circles the same size}

@@

Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supporting futures. In their queue, operations  return future objects instead of responses. Later when the response is needed it can be evaluated from the future object. They also define a weaker linearizabilty condition such that each operation can be linearized between its invocation and when its future is evaluated. Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue. They use some properties of the size of the queue before and after a batch which is similar to a part of our work. Their queue is not wait-free, in fact if the batch sizes are 1 then the queue is like MS-queue.

Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses the fast-path slow-path methodology. Their work is based on a circular queue using bounded memory. When a process wishes to do an enqueue or a dequeue, it starts two paths. One is faster but its running time is not bounded and the other one is slower but has a time bound. They show that these two paths do not effect each other and the queue remains consistent. If a process makes no progress other processes help its slow path to finish. The helping phase suffers from CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help. Because of this, the worst-case complexity cannot be better than $\Omega(p)$.

@@

David introduced the first sublinear concurrent queue \cite{DBLP:conf/wdag/David04}. Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinite memory.
The author states that to reduce memory usage to be bounded the time per operation increases to linear.

@@

\subsection{Universal Constructions and other Poly-log Time Data Structures}
A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object, introduced by Herlihy~\cite{10.1145/114005.102808}. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst case step complexity, using atomic access to $\Omega(p \log p)$ bit words. Chandra, Jayanti and Tan  introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses using reasonably sized words~\cite{DBLP:conf/podc/ChandraJT98}. Their universal construction cannot be used to create any data structure, but a set of objects called closed objects (a queue is not a closed object). We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions on page 13.

@@

It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
Instead we use an implicit representation of the operations propagated together. Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root. It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}). Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instruction on fixed-size objects in each \tt{Refresh}. To do that, we introduce \it{block}s. A block stores information about the operations propagated by a \nf{Refresh}. It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure. See Figure~\ref{fig:block} for an example. A node stores an array of blocks of operations propagated up to it. A propagate step  aggregates the new blocks in children into a new block and puts it in the parent's blocks. We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.  In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently. Thus, there are at most $p$ operations in a block. Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.

@@

\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
 [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
 [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
            [.{\begin{tabular}{||c|c||c||c|c||}  \hline (2,1) & (5,0) & (4,2) & (3,4) & (6,3) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
\end{small}
\end{tikzpicture}
\end{center}
\caption{\label{fig:block} Using blocks to represent operations. Blocks between two lines $||$ are propagated together to the parent. Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively. For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child. The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child. (5,3) is superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
\end{figure}

So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root. With this agreement on linearization ordering we can design a universal construction; for a given object we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$. However, this approach is not enough for an efficient queue.
We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization. We explain how to implement (1) and (2) in poly-logarithmic steps.

After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblocks of $B$ in the descendent of the root that contain that $i$th operation. When we reach a block in a leaf, the operation is explicitly stored there. To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block. We also store pointers to determine the range of subblocks of a block to make the binary search faster. In each block, we store the prefix sum of operations from the left child and from the right child. Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figure \ref{fig::pointer}). We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.

To compute the rank in the root of an operation in the leaf, we need to find the superblock of the block that operation is in. After a block is installed in a node we store the approximate index of its superblock in it to make this faster.

@@

In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue. In the general case a dequeue returns \nf{null} if and only if the size of the queue after the previous operation is 0. We refer to such a dequeue as a \it{null dequeue}. If the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue. Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues before the operation. So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.

In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific rank.  We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations. Consider the following operations, where operations in a cell are concurrent.
\begin{table}[hbtp]
\centering
\begin{small}
\begin{tabular}{c|c|c|c}
    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
  \end{tabular}
\end{small}
\end{table}

\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
Now, we claimed that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}. We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.

@@

Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is the number of non-null \nf{Dequeue}s in  blocks $1,2,...,b-1 + $ index of the \nf{Dequeue} in $B$\textrm{'s} \nf{Dequeue}s, if $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$. Otherwise the response would be \nf{null}.

\subsection{Details of the Implementation}
Pseudocode for the queue implementation is given in Section \ref{algQ}. It uses the following two types of objects.
\paragraph{\tt{Node}} 
 In each \texttt{Node} we store pointers to its parent and its left and right child, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.

>>>>

\begin{abstract}
In this work, we introduce a novel linearizable wait-free queue implementation.
Linearizability and lock-freedom are standard requirements for designing shared data structures. To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, the CAS Retry Problem. We show that our algorithm avoids this problem and has a worst-case running time better than prior lock-free queues. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
\end{abstract}

\tableofcontents
\pagebreak
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure to work with. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node, and then $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends the element \texttt{e} to the sequence stored. \texttt{Dequeue} removes and returns the first element in the queue. If the queue is empty, it returns \texttt{null}. In a concurrent version of the queue data structure, operations do not happen one at a time. The question that may arise is ``What properties matter for the implementation of a shared data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes. Our model is an asynchronous shared-memory distributed system of $p$ processes. Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy on how powerful objects are to solve the consensus problem among processes. Objects like LL/SC, CAS can be used to reach a consensus among any number of processors. A CAS object provides an atomic Compare \& Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}, otherwise it returns \nf{false}.  In this work, we use Compare \& Swap (\nf{CAS}) objects to synchronize among processes.

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/bad}
  \caption{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable. \texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occur before termination of \texttt{Enqueue(A)} and return \nf{null} or can occur after termination of \texttt{Enqueue(A)} and return \texttt{A}.}
\end{figure}

There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom. An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps. We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates. A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm, there might be an operation that takes infinitely many steps but never terminates.

We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps. We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized. Processes can do queries to get information about the linearization stored in the root. We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated after at most 2 \nf{CAS} invocations. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized. \nf{CAS} instructions cost more than other instructions for the processor. Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case. We use unbounded memory space in our queue and present a way to reduce the memory size to the total number of operations. Furthermore, we do not address garbage collection but we describe a way we think it can be handled.

\paragraph{Thesis outline} The rest of the thesis is organized as follows. Chapter 2 gives an outline of related work done in the area and the motivation for our implementation. Previous lock-free queues and their common problem is presented in Section 2.1. In Section 2.2, we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects. Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.

In Chapter 3 we introduce a poly-logarithmic step wait-free queue. In Section 3.1 we give a high-level description of our implementation. We also discuss the motivation and requirements in our design to achieve poly-log time. Section 3.2 contains the algorithm itself.

@@

Chapter 5 is devoted to performance analysis for our implementation. We analyze the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue. In the end, we prove our queue is wait-free.

@@

\pagebreak
\section{Related Work}
In this section, we look at previous lock-free queues. In the worst-case, step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or dequeue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.4]{pics/msqueue}
  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. In the first diagram, the first element has been dequeued. Red arrows show new pointers and gray dashed arrows show the old pointers.}
\end{figure}

@@

Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue. Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket. Each basket has a time interval in which all its nodes' enqueue operations overlap. Since the operations in a basket are concurrent, we can order them in any way. Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations. However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption{Baskets queue idea. At a time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}. To order the operations in a basket, the mechanism in the process algorithm is to \texttt{CAS} again. The successful process will be the next one in the basket, and so on.}
\end{figure}
% \bf{TODO: make small circles the same size}

@@

Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supporting futures. In their queue, operations  return future objects instead of responses. Later when the response is needed, it can be evaluated from the future object. They also define a weaker linearizability condition such that each operation can be linearized between its invocation and when its future is evaluated. Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue. They use some properties of the queue size before and after a batch, similar to a part of our work. Their queue is not wait-free, in fact if the batch sizes are 1, then the queue is like MS-queue.

Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses the fast-path slow-path methodology. Their work is based on a circular queue using bounded memory. When a process wishes to do an enqueue or a dequeue, it starts two paths. One is faster but its running time is not bounded and the other one is slower but has a time-bound. They show that these two paths do not affect each other and the queue remains consistent. If a process makes no progress, other processes help its slow path to finish. The helping phase suffers from the CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help. Because of this, the worst-case complexity cannot be better than $\Omega(p)$.

@@

David introduced the first sublinear concurrent queue \cite{DBLP:conf/wdag/David04}. Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinite memory.
The author states that to reduce memory usage to be bounded, the time per operation increases linearly.

@@

\subsection{Universal Constructions and other Poly-log Time Data Structures}
A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object, introduced by Herlihy~\cite{10.1145/114005.102808}. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$ bit words. Chandra, Jayanti and Tan  introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses using reasonably sized words~\cite{DBLP:conf/podc/ChandraJT98}. Their universal construction cannot be used to create any data structure, but a set of objects called closed objects (a queue is not a closed object). We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions on page 13.

@@

It is not efficient to explicitly store the sequence of operations in each node because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
Instead we use an implicit representation of the operations propagated together. Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root. It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}). Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instruction on fixed-size objects in each \tt{Refresh}. To do that, we introduce \it{block}s. A block stores information about the operations propagated by a \nf{Refresh}. It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure. See Figure~\ref{fig:block} for an example. A node stores an array of blocks of operations propagated up to it. A propagate step  aggregates the new blocks in children into a new block and puts it in the parent's blocks. We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.  In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently. Thus, there are at most $p$ operations in a block. Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way, if we know the number of operations from the left child and the number of operations from the right child in a block, we have a complete ordering of the operations.

@@

\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
 [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
 [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
            [.{\begin{tabular}{||c|c||c||c|c||}  \hline (2,1) & (5,0) & (4,2) & (3,4) & (6,3) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
\end{small}
\end{tikzpicture}
\end{center}
\caption{\label{fig:block} Using blocks to represent operations. Blocks between two lines $||$ are propagated together to the parent. Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively. For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child. The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child. (5,3) is the superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
\end{figure}

So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root. With this agreement on linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then return the response for $op$. However, this approach is not enough for an efficient queue.
We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization. We explain how to implement (1) and (2) in poly-logarithmic steps.

After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in the descendent of the root that contains that $i$th operation. When we reach a block in a leaf, the operation is explicitly stored there. To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block. We also store pointers to determine the range of subblocks of a block to make the binary search faster. In each block, we store the prefix sum of operations from the left child and from the right child. Moreover, for each block, we store two pointers to its last left and right subblock (see Figure \ref{fig::pointer}). We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.

To compute the rank in the root of an operation in the leaf, we need to find the superblock of the block that the operation is in. After a block is installed in a node we store the approximate index of its superblock in it to make this faster.

@@

In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue. In the general case a dequeue returns \nf{null} if and only if the queue size after the previous operation is 0. We refer to such a dequeue as a \it{null dequeue}. If the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue. Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues before the operation. So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.

In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific position.  We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations. Consider the following operations, where operations in a cell are concurrent.
\begin{table}[hbtp]
\centering
\begin{small}
\begin{tabular}{c|c|c|c}
    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
  \end{tabular}
\end{small}
\end{table}

\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
Now, we claim that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}. We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.

@@

Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is the number of non-null \nf{Dequeue}s in  blocks $1,2,...,b-1 + $ index of the \nf{Dequeue} in $B$\textrm{'s} \nf{Dequeue}s, if $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$. Otherwise, the response would be \nf{null}.

\subsection{Details of the Implementation}
Section \ref{algQ} gives the pseudocode for the queue implementation. It uses the following two types of objects.
\paragraph{\tt{Node}} 
 In each \texttt{Node} we store pointers to its parent and its left and right child, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.

\paragraph{\tt{Block}}
 The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf. If it is in a leaf, we use a \nf{LeafBlock} which stores one operation. If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$. The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the left subblocks of the block prior to $B$ ended. The right subblocks of $B$ are also defined similarly. In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}. The \texttt{sum\textsubscript{enq-left}} field is the total number of \nf{Enqueue} operations in the blocks before the last subblock of $B$ in the left child. The other fields' semantics are similar. The \nf{end\sub{left}} and \nf{end\sub{right}} field store the index of the last subblock of a block in the left and the right child, respectively. The approximate index of the superblock of non-root blocks is stored in their \nf{super} field. The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.
